Modelling with ANIMO:\\ between Fuzzy Logic and Differential Equations

### 2015-07-08 Paper.tex
Updated with latest comments. Also, removed the additional materials section (use earlier commits to get it)

\begin{abstract} % abstract <= 350 words. Headers: background, results, conclusions
\parttitle{Background}
Experimental biologists need to deal with large amounts of data on a daily basis.
This makes computational support essential in order
to reason on the dynamics of biological systems.
The software tool ANIMO (Analysis of Networks with Interactive MOdelling)
provides such computational support and allows insight into the complex networks of signalling events occurring in living cells.
Biology experts are able to exploit the power of the underlying formal methods
via a user interface specifically tailored for biological applications.
\parttitle{Results}
ANIMO models show a degree of precision intermediate between ordinary differential equations (ODEs) and fuzzy logic.
We show this by presenting ANIMO models for two case studies: \emph{Drosophila melanogaster} circadian clock, and signal
transduction events downstream of TNF$\alpha$ and EGF in HT-29 human
colon carcinoma cells.
The models were originally developed with ODEs and
fuzzy logic respectively.
\parttitle{Conclusions}
ANIMO models replicate with good precision the results of both the ODE and
fuzzy logic models. Moreover, ANIMO models require less parameters than ODEs and are more precise than
fuzzy logic. For this reason we position the modelling paradigm of ANIMO
between ODEs and fuzzy logic.
\end{abstract}

@@

\section*{Discussion}
\subsection*{Final remarks on the models}
We first described the construction of an ANIMO
model of the circadian clock in \emph{Drosophila Melanogaster}. This model
captured the dynamics of the regulatory network and led to similar
conclusions as an ODE model that had been
published previously~\cite{drosophila-ode-model}. This finding supports the use of
the series of modelling abstractions on which ANIMO is based. The biggest
difference between the construction of these models is that the model in~\cite{drosophila-ode-model}
is constructed by writing a system of mathematical equations, together
with an algorithm for simulation. In ANIMO a number of network
nodes is drawn for the molecules involved.
These nodes are then linked by directed
interactions that represent cause-and-effect relationships, with a single parameter
that defines the quantitative effect of each
interaction. This approach to construct a model is closer to biological practice.
Further contributing to an interactive modelling process
is the compositionality of the model. Each node in the network
can be disabled at any time by the user, or extra nodes can be added,
without having to change any of the existing interactions.
All this comes at the price of lower model precision: the curves representing
oscillation of protein activities in the ANIMO model are not as smooth
nor as precise as those obtained from the original ODE model (see Suppl. Fig.S6). %~\ref{suppl-fig:grafici-drosophila}).

@@

\begin{figure*}[htbp]
 \begin{center}
%   \includegraphics[width=0.9\textwidth]{Figures/2}
 \end{center}
\caption{\csentence{The spectrum of modelling methods, with the addition of ANIMO.}
The precision of ANIMO models is halfway between fuzzy logic and ODE. In addition, ANIMO
allows for an easier modelling experience thanks to a user-friendly interface based on
the widely used network modelling software Cytoscape.\label{fig:animo-spectrum}}
\end{figure*}

@@

\begin{figure*}[!htpb]
\begin{center}
%   \includegraphics[width=0.9\textwidth]{Figures/5}
\end{center}
\caption{\csentence{Signalling network downstream of TNF$\alpha$ and EGF in human colon carcinoma cells: improved model.}
{\bf(A)}
The model for the merged TNF$\alpha$ and EGF pathways
after addition of the two hypotheses (highlighted).
Hypothesis~1 assumes IL-1$\alpha$ expression to depend on AP-1 activity, which in turn requires
both c-Jun en c-Fos to be activated by JNK1 and ERK, respectively. Hypothesis~2 assumes RAS as an activator
of MEKK1. Node colours represent the activity levels $15$ minutes
after stimulation of 100 ng/ml TNF$\alpha$ + 100 ng/ml EGF.
{\bf(B)}
After the addition of the first hypothesis (activation of IL-1$\alpha$ production depending both
on JNK1 and ERK): production of IL-1$\alpha$ after stimulation with 100 ng/ml TNF$\alpha$ (series {\sf IL-1a~(TNFa)})
compared with stimulation with 100 ng/ml TGF$\alpha$ (series {\sf IL-1a~(TGFa)}) (24 hours).
{\bf(C)}
After the addition of the second hypothesis (activation of MEKK1 downstream of EGFR):
activation of JNK1 and MK2 after
stimulation with 5 ng/ml TNF$\alpha$ + 10 $\mu$g/ml C225 (2 hours).
Suppl. Sect.~D.4 %\ref{suppl:parameters-tnf-egf} 
explains how the dosage of 5 ng/ml TNF$\alpha$ was represented in the model.\\
The {\sf \_{}data} suffix identifies experimental data; all other series are computed by ANIMO.}\label{fig:large-model-complete}
\end{figure*}

@@

\appendix
\clearpage
\setcounter{figure}{0}
\setcounter{table}{0}
\setcounter{page}{1}
\onecolumn

\pagestyle{plain}

\thispagestyle{empty}
\ \\ \ \\ \ \\ \ \\ \ \\
\begin{center}
 {\huge Additional Materials}
\end{center}
\clearpage

\makeatletter

% Copied from the LaTeX sources
\def\addcontentsline#1#2#3{%
  \addtocontents{#1}{\protect\contentsline{#2}{#3}{\thepage}}}
\long\def\addtocontents#1#2{%
  \protected@write\@auxout
    {\let\label\@gobble \let\index\@gobble \let\glossary\@gobble}%
    {\string\@writefile{#1}{#2}}}
\titlecontents{section} % set formatting for \section -
                        % \subsection must be formatted separately
[2.em]                 % adjust left margin
{\bf}             % font formatting
{\contentslabel{2.em}} % section label and offset
{\hspace*{-2.em}}
{\titlerule*[1pc]{}\contentspage}
\titlecontents{subsection} % set formatting for \section -
                        % \subsection must be formatted separately
[4.5em]                 % adjust left margin
{\rmfamily}             % font formatting
{\contentslabel{2.5em}} % section label and offset
{\hspace*{2.5em}}
{\titlerule*[1pc]{.}\contentspage}
% Copied from article.cls
%\setcounter{tocdepth}{5}
\renewcommand\tableofcontents{%
\begin{spacing}{1.3}
    \section*{\contentsname
        \@mkboth{%
           \MakeUppercase\contentsname}{\MakeUppercase\contentsname}}%
    \@starttoc{toc}%
\end{spacing}
}

\makeatother

\startcontents[appendices]
\stopcontents[appendices]

\addtocontents{toc}{\protect\setcounter{tocdepth}{5}}
\thispagestyle{empty}
%\tableofcontents
\renewcommand\contentsname{Table of Contents}
\section*{Table of Contents}
\ \\ \ \\
\printcontents[appendices]{ }{1}{}
\clearpage

\resumecontents[appendices]

\setcounter{page}{1}
\setcounter{section}{0}

\renewcommand\figurename{Figure}
\renewcommand*\thefigure{S\arabic{figure}}
\renewcommand*\thetable{S\arabic{table}}

\def\ta{TA}
\def\tas{TA}

\clearpage
\section{Naming conventions}\label{suppl-sec:names}
Table~\ref{suppl-tab:names} explains the abbreviations used in the paper.

\begin{center}
\begin{longtable}{lll}
\caption{Explanation of the abbreviated names referring to molecular species
in the main text.}\label{suppl-tab:names} \\
\hline
\noalign{\vskip 2mm} {\bfseries Abbreviation} & {\bfseries Full name} & {\bfseries UniProt ID}\\[2mm]
\hline
\endfirsthead

\multicolumn{3}{c}%
{\scriptsize{\bfseries \tablename\ \thetable{}} \--\ continued from previous page} \\
\hline
\noalign{\vskip 2mm} {\bfseries Abbreviation} & {\bfseries Full name} & {\bfseries UniProt ID}\\[2mm]
\hline
\endhead

\hline \multicolumn{3}{r}{{Continued on next page}} \\ %\hline
\endfoot

\hline
\endlastfoot

\noalign{\vskip 2mm}  Akt & protein kinase B & P31749\\
\noalign{\vskip 2mm}  AP-1 & activator protein 1 & heterodimer \\
                            & & of c-Jun and c-Fos\\
\noalign{\vskip 2mm}  Casp3 & caspase 3 & P42574\\
\noalign{\vskip 2mm}  Casp8 & caspase 8 & Q14790\\
\noalign{\vskip 2mm}  c-Fos & proto-oncogene-protein c-fos & P01100\\
\noalign{\vskip 2mm}  c-Jun & Jun activation domain binding protein & P05412\\
\noalign{\vskip 2mm}  CLK & clock & O61735\\
\noalign{\vskip 2mm}  CRY & cryptochrome & O77059\\
\noalign{\vskip 2mm}  CWO & clockwork orange & Q9VGZ5\\
\noalign{\vskip 2mm}  CYC & cycle & O61734\\
\noalign{\vskip 2mm}  CYC/CLK & cycle-clock complex & \\
\noalign{\vskip 2mm}  DBT & double-time kinase & O76324\\
\noalign{\vskip 2mm}  DISC1 & death-inducing signaling complex 1 & \\
\noalign{\vskip 2mm}  DISC2 & death-inducing signaling complex 2 & \\
\noalign{\vskip 2mm}  EGF & epidermal growth factor & P01133\\
\noalign{\vskip 2mm}  EGFR & EGF receptor & P00533\\
\noalign{\vskip 2mm}  ERK & extracellular regulated kinase & P27361\\
\noalign{\vskip 2mm}  FKHR & forkhead box protein O1 & Q12778\\
\noalign{\vskip 2mm}  IKK & inhibitor of nuclear factor kappa-B kinase & O14920\\
\noalign{\vskip 2mm}  IL-1a & interleukin 1 $\alpha$ & P01583\\
\noalign{\vskip 2mm}  IL-1R & interleukin 1 receptor & P14778\\
\noalign{\vskip 2mm}  IL-1ra & interleukin 1 receptor antagonist & P18510\\
\noalign{\vskip 2mm}  IRS1 (S) & insulin receptor substrate 1 (Serine 636) & P35568\\
\noalign{\vskip 2mm}  IRS1 (Y) & insulin receptor substrate 1 (Tyrosine 896) & P35568\\
\noalign{\vskip 2mm}  JNK1 & c-Jun N-terminal kinase 1 & P45983\\
\noalign{\vskip 2mm}  MEK & MAPK ERK kinase & Q02750\\
\noalign{\vskip 2mm}  MEKK1 & MAPK/ERK kinase kinase 1 & Q13233\\
\noalign{\vskip 2mm}  MK2 & mitogen-activated protein kinase-activated\\
			  & protein kinase 2 & P49137\\
\noalign{\vskip 2mm}  MKK3/6 & dual specificity mitogen-activated\\
			     & protein kinase kinase 3/6 & P46734 / P52564\\
\noalign{\vskip 2mm}  MKK4/7 & dual specificity mitogen-activated\\
			     & protein kinase kinase 4/7 & P45985 / O14733\\
\noalign{\vskip 2mm}  NF-kB & nuclear factor kappa-B & P19838\\
\noalign{\vskip 2mm}  p38 & mitogen-activated protein kinase p38 & Q16539\\
\noalign{\vskip 2mm}  PDP1 & par-domain protein 1 & Q9TVY0\\
\noalign{\vskip 2mm}  PER & period & P07663\\
\noalign{\vskip 2mm}  PER/TIM-p & phosphorylated period-timeless complex & \\
\noalign{\vskip 2mm}  RAF & Raf & P04049\\
\noalign{\vskip 2mm}  RAS & Ras GTPase-activating protein & P01112\\
\noalign{\vskip 2mm}  TGF$\alpha$ & transforming growth factor $\alpha$ & P01135\\
\noalign{\vskip 2mm}  TNF$\alpha$ & tumor necrosis factor-$\alpha$ & P01375\\
\noalign{\vskip 2mm}  TNFR & TNF receptor & P19438\\
\noalign{\vskip 2mm}  TIM & timeless & P49021\\
\noalign{\vskip 2mm}  VRI & vrille & O18660\\[2mm]
\hline
\end{longtable}
\end{center}

\clearpage
\section{Normalizing experimental data for use with ANIMO}\label{suppl-sec:normalization}
Document S1 in the Supplemental Data of~\cite{pathway-autocrine} contains
three tables, named {\sf Replicates}, {\sf Averages} and {\sf DPLSR dataset}.
The data we use to compare the results computed with ANIMO are based on the
values from the {\sf Averages} table.
In particular, we compute activity data by performing a normalization on a $0\dots 100$ scale using this formula
$$
v_{\mbox{\scriptsize norm}} = \frac{v}{v_{\mbox{\scriptsize max}}} \times 100
$$
where $v$ is the datum to be normalized taken from the column $v$ in the {\sf Averages} table,
$v_{\mbox{\scriptsize norm}}$ is the normalized value and
$v_{\mbox{\scriptsize max}}$ is the maximum value over the whole column $v$ in the {\sf Replicates} table.
For each series, we compute also the standard deviation using the triplicate measurements
present in table {\sf Replicates}. The standard deviation is also normalized using the formula presented for the average.
Each table produced with this process contains a subset of the columns from the {\sf Averages} table,
and refers to one treatment condition only. A column with time references is added to the table in first position.
Finally, a column named \emph{Number\_of\_levels} containing only the value $100$ (see the instructions
in the ANIMO manual available online at \url{http://fmt.cs.utwente.nl/tools/animo/content/Manual.pdf}) is added at the rightmost position.
The tables are all exported in .csv format to be used with ANIMO, and are included in the \emph{Model\_and\_data.zip}
file in the additional materials of the present work.

\clearpage
\section{ANIMO and Timed Automata}\label{suppl-sec:animo-ta}
\subsection{Modelling abstractions}\label{subsec:abstractions}

In living cells, cascades of chemical and physical interactions enable propagation of signals through molecular networks.
In this process, the activity of upstream molecules induces a change in the
concentration or activity of downstream molecules. For many reactions, the values of the kinetic parameters
are unknown or difficult to collect. This lack of knowledge hampers the feasibility
of computational models that describe molecular networks in fine mechanistic detail, especially for larger networks.
As a solution to this problem, we propose the construction of models at a higher level of abstraction,
thereby reducing the number of parameters involved. In choosing a suitable abstraction level, it is important to
retain enough descriptive power to give a meaningful formal description of the topology and the
associated dynamic behaviour of biological networks.

As a first abstraction in ANIMO models, the active and inactive forms of each network component
are represented together by a single node in the network.
Each of these nodes is characterized by its \emph{activity level}, which
represents the fraction of active molecules of that molecular species. When a molecule is known to be
constitutively active, changes in concentrations of that molecule are treated as changes in its activity level.
Activity levels are discretized into integer variables with a user-defined granularity, ranging from
Boolean (2 levels) to near-continuous (100 levels).

Detailed biochemical reaction mechanisms are abstracted to \emph{interactions}, which can
represent either activations ($\rightarrow$) or inhibitions ($\dashv$\hspace{0.1em}).
This aggregation of elementary reactions into single interaction steps reduces the number of kinetic
parameters involved, while preserving cause-and-effect relationships.
For example, consider a reaction in which enzyme $E$ phosphorylates and activates substrate $S$,
transferring a phosphate group from a molecule of ATP to a molecule of $S$. Biochemically, this reaction
can be represented as
$$
\mbox{\it E} + \mbox{\it S} + \mbox{\it ATP} \rightleftarrows \mbox{\it ES} + \mbox{\it ATP} \rightarrow \mbox{\it ES}^{\mbox{\scriptsize \it P}} + \mbox{\it ADP} \rightleftarrows \mbox{\it E} + \mbox{\it S}^{\mbox{\scriptsize\it P}} + \mbox{\it ADP},
$$
with conservation condition $\mbox{\it S} + \mbox{\it S}^{\mbox{\scriptsize\it P}} = \mbox{constant}$ and $\mbox{\it ATP} + \mbox{\it ADP} = \mbox{constant}$.\\
Under the assumption of ATP constantly being replenished by the cell, this reaction is abstracted in ANIMO to the corresponding interaction
$$
\mbox{\it E} \rightarrow \mbox{\it S}.
$$
Each occurrence of the interaction $E \rightarrow S$ will increase the activity level of $S$ by one discrete step.
Since the activity level is defined as the active fraction of a molecular species, an increase in the active fraction
implies a decrease in the inactive fraction. Hence, the original conservation condition is automatically
satisfied.
The interaction rate, $R$, depends on the activity levels of the reactants involved and on a single kinetic
parameter $k$ that is set by the user.
The three available interaction scenarios can be interpreted as abstracted kinetic rate laws:
\begin{enumerate}
  \item $R = k \times [E]$: the interaction rate depends only on the activity level of the upstream node.
  \item $R = k \times [E] \times [1 - S]$ (activations) or $R = k \times [E] \times [S]$ (inhibitions): the rate
  depends on the activity levels of both the upstream and downstream participants. Activations depend on the
  presence of inactive substrate, \emph{[1 - S]}, whereas inhibitions depend on the level of active substrate,
  \emph{[S]}.
  \item $R = k \times [E_1] \times [E_2]$: this scenario can be used when the activation or inhibition
  of a downstream node depends on the simultaneous activity of two upstream nodes. This scenario is comparable to an
  \emph{AND-gate} in Boolean logic.
\end{enumerate}
As we have shown in the Results Section and in~\cite{animo-ieee,animo-gene,animo-syncop}, the abstraction described here preserves ample
expressivity to capture the dynamic behaviour of a biological network.

\subsection{ANIMO}
The modelling approach described in Section~\ref{subsec:abstractions} and in the Methods Section
is implemented in the
software tool ANIMO (Analysis of Networks with Interactive MOdelling, \cite{animo-ieee}) as
a plug-in to the network visualization tool Cytoscape~\cite{cytoscape}. The visual interface of Cytoscape
makes the construction, expansion and rewiring of a network topology a fast and user-friendly process.

When a new node is added to the network, it has to be initialized with the number of activity levels and
its initial activity.
For each interaction, a scenario needs to be selected, together with the corresponding kinetic parameter and
the interaction type: activation or inhibition. All settings can be readily adapted by double clicking, or via a
table of nodes or interactions.

ANIMO automatically translates the user input to a \tas\ model, which is then simulated with the model
checking tool UPPAAL~\cite{uppaal}. The results are subsequently parsed and translated to a graph that shows
the dynamic behaviour of nodes in the network.
A schematic overview of this process is given in Figure~\ref{fig:animo-sim-workflow}.
No training or prior knowledge on the use of \tas\ or UPPAAL is needed in order to benefit from ANIMO.
Nevertheless, the \tas\ model and the model checking process in UPPAAL can be accessed when desired by the user.

\begin{figure}[htb]
\begin{center}
\begin{minipage}{\textwidth}
\begin{center}
  \includegraphics[width=0.9\textwidth]{Figures/S1}
\caption{\csentence{The passages intercurring between the press of the \emph{Analyse network} button
and the generation of an activity level graph.}
The model represented in the Cytoscape visual interface is automatically translated by ANIMO
into a network of Timed Automata, which is then simulated using UPPAAL.
A simulation produces, for each selected node, a series of pairs {\sf ($t$, $a$)},
with $t$ a time instant along the simulation and $a$ an activity level. These data are
then parsed and represented as a graph, which is shown in ANIMO's user interface.}\label{fig:animo-sim-workflow}
\end{center}
\end{minipage}
\end{center}
\end{figure}

The dynamic behaviour of a model can be interactively explored by
moving a time slider underneath the graph to highlight time points in a simulation. In the network view,
each node will be coloured according to its activity level at the selected time point.
Experimental data can be compared to the model by importing and superposing these data
upon an output graph from the model (all the graphs in this paper were made directly in ANIMO).
The ANIMO user workflow and the
features described above are illustrated in a video on ANIMO's web page \url{http://fmt.cs.utwente.nl/tools/animo}.

\subsection{Timed Automata model}
The Timed Automata (\tas) model underlying an ANIMO network is generated whenever an analysis is requested by the user.
Starting from the network represented in the Cytoscape-based user interface, ANIMO automatically generates a \tas\ model
to be used with UPPAAL. The analysis result is then parsed and properly presented to the user, for example
as a graph of reactant activity levels. This workflow is described in Figure~\ref{fig:animo-sim-workflow}.

Each \tas\ model generated by ANIMO contains one automaton for each interaction (activation or inhibition) in the network.
A \ta\ representing an interaction performs a cyclic series of steps, continuously updating
the target of the interaction it represents, and adapting the timing of the next update according to
the user-defined dynamics. Synchronizations between different automata occur when the activity level of a network component (e.g. ERK)
changes: this allows the automata depending on that component to update their time settings.

The abstract behaviour of the interaction $\mbox{MEK} \rightarrow \mbox{ERK}$ in the \tas\ model used in ANIMO is described in Figure~\ref{fig:ta-diagram}.
There, the activity levels of MEK and ERK are represented by variables called, respectively, $\mbox{MEK}_{\mbox{\scriptsize activity}}$
and $\mbox{ERK}_{\mbox{\scriptsize activity}}$. A more detailed description of the \tas\ model underlying ANIMO was
presented in the IEEE Journal of Biomedical and Health Informatics~\cite{animo-ieee}.

\definecolor{lowActivity}{rgb}{0.8,0,0}
\definecolor{highActivity}{rgb}{0,0.8,0}%{1,0.8,0}
\newsavebox{\mysquareLow}
\savebox{\mysquareLow}{%
  \raisebox{-0.08em}{%
    \textcolor{black}{%
      \rule{.7em}{.7em}%
    }%
    \hspace{-.65em}%
    \raisebox{.05em}{%
      \textcolor{lowActivity}{%
	\rule{.6em}{.6em}%
      }%
    }%
  }%
}
\newsavebox{\mysquareHigh}
\savebox{\mysquareHigh}{%
  \raisebox{-0.08em}{%
    \textcolor{black}{%
      \rule{.7em}{.7em}%
    }%
    \hspace{-.65em}%
    \raisebox{.05em}{%
      \textcolor{highActivity}{%
	\rule{.6em}{.6em}%
      }%
    }%
  }%
}

\begin{figure}[!ht]
\begin{minipage}{\textwidth}
\centering
  \includegraphics[width=0.5\textwidth]{Figures/S2}
\caption{\csentence{Schematic overview of the steps taken during a simulation run by a Timed Automaton modelling an interaction $R$ that
increases $\mbox{ERK}_{\mbox{\scriptsize activity}}$ and depends only on $\mbox{MEK}_{\mbox{\scriptsize activity}}$.}
In this example, MEK is represented with a granularity of 10 activity levels.
After resetting the internal clock {\sf t}, the automaton sets the time constraints for the interaction.
$\mbox{MEK}_{\mbox{\scriptsize activity}}$ is used as the index inside the time
tables $R_{\mbox{\scriptsize lowerBound}}$ and $R_{\mbox{\scriptsize upperBound}}$, which contain pre-computed lower- and upper-bounds
for the interaction timing.
Once the bounds have been identified, %the internal clock {\sf t} begins counting.
$R$ can occur when {\sf t} reaches a value
inside the continuous time interval~$[\,\usebox{\mysquareLow}, \usebox{\mysquareHigh}\,]$. When it occurs, $R$ increases the value of
$\mbox{ERK}_{\mbox{\scriptsize activity}}$ by $1$. All interactions that depend on
$\mbox{ERK}_{\mbox{\scriptsize activity}}$ are notified of the change (via a synchronization on a specific channel),
so that the associated time bounds are updated accordingly.
After resetting the clock {\sf t}, the process can restart.
If $\mbox{MEK}_{\mbox{\scriptsize activity}}$ was changed by another automaton before the occurrence of $R$,
the time bounds are updated according to the new activity level of MEK.}\label{fig:ta-diagram}
\end{minipage}
\end{figure}

\subsection{Granularity of an ANIMO network node}
Figure~\ref{fig:levels} shows the differences between different choices for the
number of levels of a node. This allows to adapt a model to the quality of experimental data.

\begin{figure}[htpb]
\begin{minipage}{\textwidth}
\centering
  \includegraphics[width=\textwidth]{Figures/S3}
\caption{\csentence{Comparing different reactant granularity settings.}
{\bf(A)} 2 levels,
{\bf(B)} 10 levels,
{\bf (C)} 50 levels,
{\bf (D)} 100 levels.
The {\sf JNK1} series is computed from the model presented in Figure~\ref{fig:large-model-complete},
considering 100 ng/ml TNF$\alpha$ as treatment condition over a period of 60 minutes.}\label{fig:levels}
\end{minipage}
\end{figure}

\clearpage
\section{Additional notes}

\subsection{ANIMO model of the Drosophila circadian clock}
The model built in ANIMO to represent the circadian clock of \emph{Drosophila Melanogaster}
can be seen in Figure~\ref{fig:drosophila-model}. Its topology is based on the one presented
in~\cite{drosophila-ode-model}, while the parameters were set by hand in order to obtain
an oscillatory behavior.

\begin{figure*}[!htpb]
\begin{center}
  \includegraphics[width=0.75\textwidth]{Figures/S4}
\end{center}
\caption{\csentence{ANIMO model of the circadian clock in \emph{Drosophila Melanogaster}.}
Auto-regulatory negative feedback loops are present on each of the nodes of
the network, following the original model in~\cite{drosophila-ode-model}. These feedback loops ensure that
protein levels decrease over time when activating inputs are absent. The feedback loops are not represented here
for cosmetic reasons only.
Naming conventions follow the same rules
as in the original model, with lower-case names representing mRNA, and upper-case names indicating proteins.
}\label{fig:drosophila-model}
\end{figure*}

\subsection{Simulating the day-night cycle}\label{suppl:repressilator}
The model presented in Figure~\ref{fig:drosophila-model} contains a node
labelled {\sf Day/Night}. That node abstracts our representation
of the cyclic alternation of day and night, which causes the variations
in cryptochrome ({\sf cry}): these oscillations allow the network
to synchronize to a time zone. Note that the network oscillates
also when the node {\sf cry} is not included in the model.

The alternation between day and night is represented in our model with a
repressilator-like~\cite{repressilator} subnetwork, as can be seen in Figure~\ref{fig:repressilator}.
In the model in~\cite{drosophila-ode-model} a specific function
was introduced in the equations to approximate the experimental data from~\cite{drosophila-cry-data}.

\def\reprGraph{\includegraphics[scale=0.15]{images/drosophila_model_repressilator}}
\newlength\reprGraphHeight
\setlength\reprGraphHeight{\heightof{\reprGraph}}
\begin{figure}[!htb]
\begin{minipage}{\textwidth}
  \centering
  \includegraphics[width=0.6\textwidth]{Figures/S5}
\caption{\csentence{Modelling the dany/night cycle in ANIMO.}{\bf (A)} The repressilator-like subnetwork used to represent the alternation
between day and night that cause the oscillations in {\sf CRY} concentrations in the
network modelled in the Results Section.
{\bf (B)} A graph plotting the oscillations in {\sf CRY} along
a period of three days.}\label{fig:repressilator}
\end{minipage}
\end{figure}

\subsection{Comparing the models of the Drosophila circadian clock}\label{suppl-sec:animo-drosophila}
We compared the simulation results from the ANIMO model presented in Figure~\ref{fig:drosophila-model}
with the ODE model described in~\cite{drosophila-ode-model}. The raw data coming from the
two models were aligned to have a roughly close initial point, and all amplitudes were normalized
following the procedure described in Suppl. Sect.~\ref{suppl-sec:normalization}. The results
of this comparison can be seen in Figure~\ref{suppl-fig:grafici-drosophila}.

\def\drosophilaGraphScale{0.0685}
\renewcommand*\thesubfigure{}
\begin{figure}[!htb]
\begin{minipage}{\textwidth}
  \centering
  \pgfuseimage{FigS6}
\caption{\csentence{Comparing the simulation results from the ANIMO and ODE models of
\emph{Drosophila Melanogaster} circadian clock.} Blue lines ({\sf \_{}ANIMO} series)
represent data from the ANIMO model in Figure~\ref{fig:drosophila-model},
while red lines ({\sf \_{}ODE} series) represent data from the ODE model in~\cite{drosophila-ode-model}.\label{suppl-fig:grafici-drosophila}}
\end{minipage}
\end{figure}

Most of the molecules represented in the two models evolve with the same period and phase.
CLK and clk in the ANIMO model have
a small oscillation range (their values change by around 10\%\ during a simulation),
so their behaviour match the continuous model less precisely.

\subsection{Note on the parameters in the TNF$\alpha$-EGF model}\label{suppl:parameters-tnf-egf}
The parameters in the models in Figures~\ref{fig:large-model-all} and~\ref{fig:large-model-complete}
have been set by fitting the model to the experimental data for the treatment conditions with 100 ng/ml TNF$\alpha$, 
100 ng/ml EGF and 100 ng/ml TNF$\alpha$ + 100 ng/ml EGF.
In the model we have set the activity of the corresponding {\sf TNFa} and {\sf EGF} nodes either at
their maximum value or at 0, depending on the modelled treatment.
This level is a dimensionless quantity that indicates either the maximum or minimum activity level in the data set.
However, we found that in order to model the condition with 5 ng/ml TNF$\alpha$
we needed to set the activity of the {\sf TNFa} node to a value higher than 5 out of 100. 
We believe that this has to do with the fact that
100 ng/ml is a highly supra-physiological concentration of TNF$\alpha$, that will rapidly cause activation of all
receptors present. Fitting the model to this experimental condition may have resulted in slight deviations
in the parameter values. Nevertheless, the modelling results illustrate that building a model with basic
kinetic rate laws can give useful predictions over a range of concentrations.
Figure~\ref{fig:large-model-all}C was obtained using the model
in Figure~\ref{fig:large-model-all}A and setting the activity of {\sf TNFa} to 12 out of 100 (lower
values gave 0 activity for both JNK1 and MK2),
while Figure~\ref{fig:large-model-complete}C was obtained from the model in Figure~\ref{fig:large-model-complete}A
with {\sf TNFa} set at 8 out of 100.

\section{Supplementary figures}\label{sec:supplementary-figures}

\begin{figure}[htpb]
\begin{minipage}{\textwidth}
\centering
  \includegraphics[width=.7\textwidth]{Figures/S7}
\caption{\csentence{The model for the TNF$\alpha$ pathway in isolation.} Node colours represent the activity level of the
corresponding modelled reactants at time $t = 10$ minutes after a stimulation of 100 ng/ml TNF$\alpha$.}\label{fig:large-model-tnf}
\end{minipage}
\end{figure}

\begin{figure}[!tpb]
\begin{minipage}{\textwidth}
\centering
  \includegraphics[width=.7\textwidth]{Figures/S8}
\caption{\csentence{The model for the EGF pathway in isolation.} Node colours represent the
activity level of the corresponding modelled reactants at time $t = 5$ minutes after
a stimulation of 100 ng/ml EGF.}\label{fig:large-model-egf}
\end{minipage}
\end{figure}

\begin{figure}[!tpb]
\begin{minipage}{\textwidth}
\centering
  \includegraphics[width=0.9\textwidth]{Figures/S9}
\caption{\csentence{The merged model for the TNF$\alpha$-EGF pathway with the two hypotheses are highlighted.}
The first hypothesis is the dependence IL-1$\alpha$ expression on the
combined activity of ERK and JNK1. The second hypothesis assumes an as yet unidentified protein (Hyp2) to link EGFR to MEKK1.
Node colours represent initial activity levels.}\label{fig:large-model-hypotheses}
\end{minipage}
\end{figure}

>>>>

\begin{abstract} % abstract <= 350 words. Headers: background, results, conclusions
\parttitle{Background}
Experimental biologists need to deal with large amounts of data on a daily basis.
This makes computational support essential in order
to reason on the dynamics of biological systems.
The software tool ANIMO (Analysis of Networks with Interactive MOdelling)
provides such computational support and allows insight into the complex networks of signalling events occurring in living cells.
Biology experts are able to exploit the power of the underlying formal methods
via a user interface specifically tailored for biological applications. Knowledge
of the underlying formalisms is not necessary to use ANIMO.
\parttitle{Results}
ANIMO models show a degree of precision intermediate between ordinary differential equations (ODEs) and fuzzy logic.
We show this by presenting ANIMO models for two case studies: \emph{Drosophila melanogaster} circadian clock, and signal
transduction events downstream of TNF$\alpha$ and EGF in HT-29 human
colon carcinoma cells.
The models were originally developed with ODEs and
fuzzy logic, respectively.
\parttitle{Conclusions}
ANIMO models replicate with good precision the results of both the ODE and
fuzzy logic models. Moreover, ANIMO models require less parameters than ODEs and are more precise than
fuzzy logic. For this reason we position the modelling paradigm of ANIMO
between ODEs and fuzzy logic.
\end{abstract}

@@

\section*{Discussion}
\subsection*{Final remarks on the models}
We first described the construction of an ANIMO
model of the circadian clock in \emph{Drosophila Melanogaster}. This model
captured the dynamics of the regulatory network and led to similar
conclusions as an ODE model that had been
published previously~\cite{drosophila-ode-model}. This finding supports the use of
the series of modelling abstractions on which ANIMO is based. The biggest
difference between the construction of these models is that the model in~\cite{drosophila-ode-model}
is constructed by writing a system of mathematical equations, together
with an algorithm for simulation. In ANIMO a number of network
nodes is drawn for the molecules involved.
These nodes are then linked by directed
interactions that represent cause-and-effect relationships, with a single parameter
that defines the quantitative effect of each
interaction. This approach to construct a model is closer to biological practice.
Further contributing to an interactive modelling process
is the compositionality of the model. Each node in the network
can be disabled at any time by the user, or extra nodes can be added,
without having to change any of the existing interactions.
All this comes at the price of lower model precision: the curves representing
oscillation of protein activities in the ANIMO model are not
as precise as those obtained from the original ODE model (see Suppl. Fig.S6). %~\ref{suppl-fig:grafici-drosophila}).

@@

\begin{figure*}[htbp]
 \begin{center}
%   \includegraphics[width=0.9\textwidth]{Figures/2}
 \end{center}
\caption{\csentence{The spectrum of modelling methods, with the addition of ANIMO.}
The precision of ANIMO models is halfway between fuzzy logic and ODE. Compared to other
modelling tools, ANIMO
allows for an easier modelling experience thanks to a user-friendly interface based on
the widely used network modelling software Cytoscape.\label{fig:animo-spectrum}}
\end{figure*}

@@

\begin{figure*}[!htpb]
\begin{center}
%   \includegraphics[width=0.9\textwidth]{Figures/5}
\end{center}
\caption{\csentence{Signalling network downstream of TNF$\alpha$ and EGF in human colon carcinoma cells: improved model.}
{\bf(A)}
The model for the merged TNF$\alpha$ and EGF pathways
after addition of the two hypotheses (highlighted).
Hypothesis~1 assumes IL-1$\alpha$ expression to depend on AP-1 activity, which in turn requires
both c-Jun en c-Fos to be activated by JNK1 and ERK, respectively. Hypothesis~2 assumes RAS as an activator
of MEKK1. Node colours represent the activity levels $15$ minutes
after stimulation of 100 ng/ml TNF$\alpha$ + 100 ng/ml EGF.
{\bf(B)}
After the addition of the first hypothesis (activation of IL-1$\alpha$ production depending both
on JNK1 and ERK): production of IL-1$\alpha$ after stimulation with 100 ng/ml TNF$\alpha$ (series {\sf IL-1a~(TNFa)})
compared with stimulation with 100 ng/ml TGF$\alpha$ (series {\sf IL-1a~(TGFa)}) (24 hours). The {\sf IL-1a (TGFa)} series is always 0.
{\bf(C)}
After the addition of the second hypothesis (activation of MEKK1 downstream of EGFR):
activation of JNK1 and MK2 after
stimulation with 5 ng/ml TNF$\alpha$ + 10 $\mu$g/ml C225 (2 hours). The {\sf JNK1} series is always 0.
Suppl. Sect.~D.4 %\ref{suppl:parameters-tnf-egf} 
explains how the dosage of 5 ng/ml TNF$\alpha$ was represented in the model.\\
The {\sf \_{}data} suffix identifies experimental data; all other series are computed by ANIMO.}\label{fig:large-model-complete}
\end{figure*}

@@
